Creating optimal conditions for reproducible data analysis in \R with `fertile'\protect\thanks{The authors gratefully acknowledge contributions from Hadley Wickham, Jenny Bryan, Greg Wilson, Edgar Ruiz, and other members of the \pkg{tidyverse} team.}

### 2020-08-09 STAT-paper-template/APA/fertile.Rnw
condensed spacing

\usepackage{listings}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\graphicspath{{images/}}

@@

However, while the necessity of reproducibility is clear, there are significant behavioral and technical challenges that impede its widespread implementation, and no clear consensus on standards of what constitutes reproducibility in published research (\cite{peng2009bio}). Not only are the components of reproducible research up for discussion (e.g., need the software be open source?), but the corresponding recommendations for ensuring reproducibility also vary (e.g., should raw and processed data files be in separate directories?).

Existing attempts to address reproducibility in data science are often either too generalized---resulting in shallow and vague recommendations that are challenging to implement---or too specific---approaching one aspect of reproducibility well but doing so in a highly technical way that fails to capture the bigger picture and creates challenges for inexperienced users.

@@

Publications on reproducibility can be found in all areas of scientific research. However, as \cite{Goodman341ps12} argue, the language and conceptual framework of research reproducibility varies significantly across the sciences, and there are no clear standards on reproducibility agreed upon by the scientific community as a whole.

We consider recommendations from a variety of fields, and determine the key aspects of reproducibility faced by scientists in different disciplines.

\cite{kitzes2017practice} present a collection of case studies on reproducibility practices from across the data-intensive sciences, illustrating a variety of recommendations and techniques for achieving reproducibility. Although the book does not come to a consensus on the exact standards of reproducibility that should be followed, several common trends and principles emerge from their case studies:

\begin{enumerate}
\item Use clear separation, labeling, and documentation.
\item Automate processes when possible.
\item Design the data analysis workflow as a sequence of small steps glued together, with
outputs from one step serving as inputs into the next. This is a common suggestion within the computing community, originating as part of the Unix philosophy (\cite{unix}).
\end{enumerate}

\cite{cooper2017guide} focuses on data analysis in \R and identifies a similar list of important reproducibility components and reiterates the need for clearly labeled, well-documented, and well-separated files. In addition, they recommend publishing a list of dependencies and using version control.
\cite{broman} reiterates the need for clear naming and file separation while sharing several additional suggestions: keep the project contained in one directory, use relative paths, and include a \cmd{README}.

@@

When considered in combination, these sources provide a well-rounded picture of the components important to research reproducibility. Using this literature as a guideline, we identify several key features of reproducible work. These recommendations are opinionated---due to the lack of agreement on which components of reproducibility are most important, we select those that are mentioned most often, as well as some that are mentioned less but that we view as important.

\begin{enumerate}
\item A well-designed file structure:
  \begin{itemize}
  \item Separate folders for different file types.
  \item No extraneous files.
  \item Minimal clutter.
  \end{itemize}
\item Good documentation:
  \begin{itemize}
  \item Files are clearly named, preferably in a way where the order in which they should be run is clear.
  \item A README is present.
  \item Dependencies are noted.
  \end{itemize}
\item Reproducible file paths:
  \begin{itemize}
  \item No absolute paths, or paths leading to locations outside of a project's directory, are used in code---only portable (relative) paths.
  \end{itemize}
\item Randomness is accounted for:
  \begin{itemize}
  \item If randomness is used in code, a seed must also be set.
  \end{itemize}
\item Readable, styled code:
  \begin{itemize}
  \item Code should be written in a coherent style. Code that conforms to a style guide or is written in a consistent dialect is easier to read (\cite{hermans2017programming}). We believe that the \pkg{tidyverse} provides the most accessible dialect of \R.
  \end{itemize}
\end{enumerate}

@@

Much of the work discussed in Section \ref{sec:papers} is highly generalized, written to be applicable to users working with a variety of statistical software programs. Because all statistical software programs operate differently, these recommendations are inherently vague and difficult to implement, particularly to new analysts who are relatively unfamiliar with their software.

Focused attempts to address reproducibility in specific certain software programs are more likely to be successful. 
We focus on \R, due to its open-source nature, accessibility, and popularity as a tool for statistical analysis.

A small body of \R packages focus on research reproducibility.

\pkg{rrtools} (\cite{R-rrtools}) addresses some of the issues discussed in \cite{marwick2018packaging} by creating a basic \R package structure for a data analysis project and implementing a basic \func{testthat::check} functionality. The \pkg{orderly} (\cite{R-orderly}) package also focuses on file structure, requiring the user to declare a desired project structure (typically a step-by-step structure, where outputs from one step are inputs into the next) at the beginning and then creating the files necessary to achieve that structure. \pkg{workflowr}'s (\cite{R-workflowr}) functionality is based around version control and making code easily available online. It works to generate a website containing time-stamped, versioned, and documented results. \pkg{checkers} (\cite{R-checkers}) allows you to create custom checks that examine different aspects of reproducibility. \pkg{packrat} (\cite{R-packrat}) is focused on dependencies, creating a packaged folder containing a project as well as all of its dependencies, so that projects dependent on lesser-used packages can be easily shared across computers. \pkg{drake} (\cite{R-drake}) analyzes workflows, skip steps where results are up to date, and provides evidence that results match the underlying code and data. Lastly, the \pkg{reproducible} (\cite{R-reproducible}) package focuses on the concept of caching: saving information so that projects can be run faster each time they are re-completed from the start.

@@

Users can access the majority of \pkg{fertile}'s retroactive features through two primary functions, \func{proj\_check} and \func{proj\_analyze}.
The \func{proj\_check} function runs fifteen different reproducibility tests, noting which ones passed, which ones failed, the reason for failure, a recommended solution, and a guide to where to look for help. These tests include: looking for a clear build chain, checking to make sure the root level of the project is clear of clutter, confirming that there are no files present that are not being directly used by or created by the code, and looking for uses of randomness that do not have a call to \func{set.seed} present. A full list is provided below:

@@

Each test can also be run individually by calling the function matching its check name:

<<individual checks, warning = FALSE>>=

has_tidy_raw_data(proj_dir)

@

@@

<<proj_analyze>>=
proj_analyze(proj_dir)
@

@@

<<read-csv, message=FALSE, include=FALSE>>=
library(purrr)
library(forcats)
library(stringr)
read_csv(fs::path(proj_dir, "mice.csv"))
@

@@

\begin{enumerate}
\item Identify an \R function that is likely to be involved in operations that may break reproducibility. Popular functions associated with only one package (e.g., \func{read\_csv} from \pkg{readr}) are ideal candidates.
\item Create a function in \pkg{fertile} with the same name that takes the same arguments (and always the dots \cmd{...}).
\item Write this new function so that it: 
\begin{enumerate}
\item Captures any arguments
\item Logs the name of the function called
\item Performs any checks on these arguments, and, finally:
\item Calls the original function with the original arguments. Except where warranted, the execution looks the same to the user as if they were calling the original function.
\end{enumerate}
\end{enumerate}
Most shims are quite simple and look something like what is shown below for \func{read\_csv}.

@@

In order to ensure that the \pkg{fertile} versions of functions ("shims") are always favored ("masked") over their original namesakes when called, \pkg{fertile} uses its own shims of the \func{library} and \func{require} functions to  manipulate the \R \func{search} path so that it is always located in the first position. In the \pkg{fertile} version of \func{library}, we detach \pkg{fertile} from the search path, load the requested package, and then re-attach \pkg{fertile}. This ensures that when a user executes a command, \R will check \pkg{fertile} for a matching function before considering other packages.

While it is possible that this shifty behavior could lead to unintended consequences, our goal is to catch a good deal of problems before they become problematic. Users can easily disable \pkg{fertile} by detaching it, or not loading it in the first place.

@@

<<path checks, error = TRUE>>=

# Path inside the directory
check_path("project_miceps")

@@

# Path outside the directory
check_path("../fertile.Rmd")

@

@@

<<is_file>>=
is_data_file(fs::path(proj_dir, "mice.csv"))
is_image_file(fs::path(proj_dir, "mice.csv"))
is_image_file(fs::path(proj_dir, "proteins_v_time.png"))
is_text_file(fs::path(proj_dir, "README.md"))
is_r_file(fs::path(proj_dir, "analysis.Rmd"))
@

@@

<<sandbox>>=
dir <- getwd()
new_dir <- sandbox(dir)

dir
fs::dir_ls(dir) %>% head(3)

new_dir
fs::dir_ls(new_dir) %>% head(3)
@

@@

\pkg{fertile} is designed to: 
\begin{enumerate}
\item Be simple enough that users with minimal \R experience can use the package without issue.
\item Increase the reproducibility of work produced by its users, and:
\item Educate its users on why their work is or is not reproducible and provide guidance on how to address any problems.
\end{enumerate}

@@

\begin{itemize}

@@

Both groups are given a short quiz on different components of reproducibility that are intended to be taught by \pkg{fertile} at both the beginning and end of the semester. Their scores are then compared to see whether one group learned more than the other group or whether their scores were essentially equivalent. Additionally, for every homework assignment submitted, the professor takes note of whether or not the project compiles successfully.

@@

Based on the results, we could determine whether \pkg{fertile} was successful at achieving its intended goals:

\begin{enumerate}

\item A lack of notable difference between the \func{experimental} and \func{control} groups in terms of the number of code-related questions asked throughout the semester would indicate that \pkg{fertile} achieved its goal of simplicity.
\item A higher average for the \func{experimental} group in terms of the number of homework assignments that compiled successfully would indicate that \pkg{fertile} was successful in increasing reproducibility.
\item A greater increase over the semester in the reproducibility quiz scores for students in the \func{experimental} group compared with the \func{control} group would indicate that \pkg{fertile} achieved its goal of educating users on reproducibility.

\end{enumerate}

Success in achieving these goals would provide valuable evidence showing \pkg{fertile}'s benefit as tool to help educators introduce reproducibility concepts in the classrom.

@@

\begin{itemize}
  \item Expanding dependency management features to include \R session information and package version numbers in addition to package names.
  \item Expanding code and documentation style features to analyze whether code has been properly commented in addition to checking for a \cmd{README} and tidy code style.
  \item Adding \cmd{make}-like functionality that can analyze an \R project structure and files and use this information to generate a Makefile. This Makefile would have information about target files and their prerequisites and would assist with making sure that re-running an analysis is done as quickly as possible by ensuring that only the necessary code and files that have been updated are run when rebuilding and re-running code.
\end{itemize}

>>>>

\usepackage{listings}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{enumitem}
\usepackage{titlesec}
\graphicspath{{images/}}
\titlespacing{\section}{0pt}{10pt plus 2pt minus 1pt}{10pt plus 2pt minus 1pt}
\titlespacing{\subsection}{0pt}{7pt plus 2pt minus 1pt}{7pt plus 2pt minus 1pt}
\titlespacing{\subsubsection}{0pt}{7pt plus 2pt minus 1pt}{7pt plus 2pt minus 1pt}

@@

However, while the necessity of reproducibility is clear, there are significant behavioral and technical challenges that impede its widespread implementation, and no clear consensus on standards of what constitutes reproducibility in published research (\cite{peng2009bio}). Not only are the components of reproducible research up for discussion (e.g., need the software be open source?), but the corresponding recommendations for ensuring reproducibility also vary (e.g., should raw and processed data files be in separate directories?).
Existing attempts to address reproducibility in data science are often either too generalized---resulting in shallow and vague recommendations that are challenging to implement---or too specific---approaching one aspect of reproducibility well but doing so in a highly technical way that fails to capture the bigger picture and creates challenges for inexperienced users.

@@

Publications on reproducibility can be found in all areas of scientific research. However, as \cite{Goodman341ps12} argue, the language and conceptual framework of research reproducibility varies significantly across the sciences, and there are no clear standards on reproducibility agreed upon by the scientific community as a whole. We consider recommendations from a variety of fields, and determine the key aspects of reproducibility faced by scientists in different disciplines.

\cite{kitzes2017practice} present a collection of case studies on reproducibility practices from across the data-intensive sciences, illustrating a variety of recommendations and techniques for achieving reproducibility. Although the book does not come to a consensus on the exact standards of reproducibility that should be followed, several common trends and principles emerge from their case studies: 1) Use clear separation, labeling, and documentation, 2) Automate processes when possible, and 3) Design the data analysis workflow as a sequence of small steps glued together, with outputs from one step serving as inputs into the next. This is a common suggestion within the computing community, originating as part of the Unix philosophy (\cite{unix}).

\cite{cooper2017guide} focuses on data analysis in \R and identifies a similar list of important reproducibility components and reiterates the need for clearly labeled, well-documented, and well-separated files. In addition, they recommend publishing a list of dependencies and using version control.

\cite{broman} reiterates the need for clear naming and file separation while sharing several additional suggestions: keep the project contained in one directory, use relative paths, and include a \cmd{README}.

@@

When considered in combination, these sources provide a well-rounded picture of the components important to research reproducibility. Using this literature as a guideline, we identify several key features of reproducible work. These recommendations are opinionated---due to the lack of agreement on which components of reproducibility are most important, we select those that are mentioned most often, as well as some that are mentioned less but that we view as important. 
\vspace{3mm}

\begin{enumerate} [nolistsep]
\item A well-designed file structure:
  \begin{itemize}[noitemsep]
  \item Separate folders for different file types.
  \item No extraneous files.
  \item Minimal clutter.
  \end{itemize}
\item Good documentation:
  \begin{itemize} [noitemsep]
  \item Files are clearly named, preferably in a way where the order in which they should be run is clear.
  \item A README is present.
  \item Dependencies are noted.
  \end{itemize}
\item Reproducible file paths:
  \begin{itemize}
  \item No absolute paths, or paths leading to locations outside of a project's directory, are used in code---only portable (relative) paths.
  \end{itemize}
\item Randomness is accounted for:
  \begin{itemize}
  \item If randomness is used in code, a seed must also be set.
  \end{itemize}
\item Readable, styled code:
  \begin{itemize}
  \item Code should be written in a coherent style. Code that conforms to a style guide or is written in a consistent dialect is easier to read (\cite{hermans2017programming}). We believe that the \pkg{tidyverse} provides the most accessible dialect of \R.
  \end{itemize}
\end{enumerate}

@@

Much of the work discussed in Section \ref{sec:papers} is highly generalized, written to be applicable to users working with a variety of statistical software programs. Because all statistical software programs operate differently, these recommendations are inherently vague and difficult to implement, particularly to new analysts who are relatively unfamiliar with their software. Focused attempts to address reproducibility in specific certain software programs are more likely to be successful. We focus on \R, due to its open-source nature, accessibility, and popularity as a tool for statistical analysis.

A small body of \R packages focus on research reproducibility. \pkg{rrtools} (\cite{R-rrtools}) addresses some of the issues discussed in \cite{marwick2018packaging} by creating a basic \R package structure for a data analysis project and implementing a basic \func{testthat::check} functionality. The \pkg{orderly} (\cite{R-orderly}) package also focuses on file structure, requiring the user to declare a desired project structure (typically a step-by-step structure, where outputs from one step are inputs into the next) at the beginning and then creating the files necessary to achieve that structure. \pkg{workflowr}'s (\cite{R-workflowr}) functionality is based around version control and making code easily available online. It works to generate a website containing time-stamped, versioned, and documented results. \pkg{checkers} (\cite{R-checkers}) allows you to create custom checks that examine different aspects of reproducibility. \pkg{packrat} (\cite{R-packrat}) is focused on dependencies, creating a packaged folder containing a project as well as all of its dependencies, so that projects dependent on lesser-used packages can be easily shared across computers. \pkg{drake} (\cite{R-drake}) analyzes workflows, skip steps where results are up to date, and provides evidence that results match the underlying code and data. Lastly, the \pkg{reproducible} (\cite{R-reproducible}) package focuses on the concept of caching: saving information so that projects can be run faster each time they are re-completed from the start.

@@

Users can access the majority of \pkg{fertile}'s retroactive features through two primary functions, \func{proj\_check} and \func{proj\_analyze}.

The \func{proj\_check} function runs fifteen different reproducibility tests, noting which ones passed, which ones failed, the reason for failure, a recommended solution, and a guide to where to look for help. These tests include: looking for a clear build chain, checking to make sure the root level of the project is clear of clutter, confirming that there are no files present that are not being directly used by or created by the code, and looking for uses of randomness that do not have a call to \func{set.seed} present. A full list is provided below:

@@

Each test can also be run individually by calling the function matching its check name.

@@

<<proj_analyze, out.width = 180>>=
proj_analyze(proj_dir)
@

@@

<<read-csv, message=FALSE, include=FALSE>>=
library(purrr)
library(forcats)
read_csv(fs::path(proj_dir, "mice.csv"))
@

@@

\begin{enumerate}[noitemsep]
\item Identify an \R function that is likely to be involved in operations that may break reproducibility. Popular functions associated with only one package (e.g., \func{read\_csv} from \pkg{readr}) are ideal candidates.
\item Create a function in \pkg{fertile} with the same name that takes the same arguments (and always the dots \cmd{...}).
\item Write this new function so that it: A) Captures any arguments, B) Logs the name of the function called, C) Performs any checks on these arguments, and D) Calls the original function with the original arguments. Except where warranted, the execution looks the same to the user as if they were calling the original function.
\end{enumerate}
Most shims are quite simple and look something like what is shown below for \func{read\_csv}.

@@

In order to ensure that the \pkg{fertile} versions of functions ("shims") are always favored ("masked") over their original namesakes when called, \pkg{fertile} uses its own shims of the \func{library} and \func{require} functions to  manipulate the \R \func{search} path so that it is always located in the first position. In the \pkg{fertile} version of \func{library}, we detach \pkg{fertile} from the search path, load the requested package, and then re-attach \pkg{fertile}. This ensures that when a user executes a command, \R will check \pkg{fertile} for a matching function before considering other packages. While it is possible that this shifty behavior could lead to unintended consequences, our goal is to catch a good deal of problems before they become problematic. Users can easily disable \pkg{fertile} by detaching it, or not loading it in the first place.

@@

<<path checks, error = TRUE, message = FALSE>>=
# Path inside the directory
check_path("project_miceps")

@@

# Path outside the directory
check_path("../fertile.Rmd")
@

@@

<<is_file>>=
is_data_file(fs::path(proj_dir, "mice.csv"))
is_image_file(fs::path(proj_dir, "proteins_v_time.png"))
is_text_file(fs::path(proj_dir, "README.md"))
is_r_file(fs::path(proj_dir, "analysis.Rmd"))
@

@@

<<sandbox>>=
proj_dir
fs::dir_ls(proj_dir) %>% head(1)
temp_dir <- sandbox(proj_dir)
temp_dir
fs::dir_ls(temp_dir) %>% head(1)
@

@@

\pkg{fertile} is designed to: 1) Be simple enough that users with minimal \R experience can use the package without issue, 2) Increase the reproducibility of work produced by its users, and 3) Educate its users on why their work is or is not reproducible and provide guidance on how to address any problems.

@@

\begin{itemize}[noitemsep]

@@

\item Both groups are given a short quiz on different components of reproducibility that are intended to be taught by \pkg{fertile} at both the beginning and end of the semester. Their scores are then compared to see whether one group learned more than the other group or whether their scores were essentially equivalent. Additionally, for every homework assignment submitted, the professor takes note of whether or not the project compiles successfully.

@@

Based on the results, we could determine whether \pkg{fertile} was successful at achieving its intended goals: 1) A lack of notable difference between the \func{experimental} and \func{control} groups in terms of the number of code-related questions asked throughout the semester would indicate that \pkg{fertile} achieved its goal of simplicity, 2) A higher average for the \func{experimental} group in terms of the number of homework assignments that compiled successfully would indicate that \pkg{fertile} was successful in increasing reproducibility, and 3) A greater increase over the semester in the reproducibility quiz scores for students in the \func{experimental} group compared with the \func{control} group would indicate that \pkg{fertile} achieved its goal of educating users on reproducibility. Success in achieving these goals would provide valuable evidence showing \pkg{fertile}'s benefit as tool to help educators introduce reproducibility concepts in the classrom.

@@

\begin{itemize} [noitemsep]
  \item Expanding dependency management features to include \R session information and package version numbers in addition to package names.
  \item Expanding code and documentation style features to analyze whether code has been properly commented in addition to checking for a \cmd{README} and tidy code style.
  \item Adding \cmd{make}-like functionality that can analyze an \R project structure and files and use this information to generate a Makefile. This Makefile would have information about target files and their prerequisites and would assist with making sure that re-running an analysis is done as quickly as possible by ensuring that only the necessary code and files that have been updated are run when rebuilding and re-running code.
\end{itemize}
